---
title: "VisNetwork: an introduction with Game of Thrones..."
author: "Faisal Samin, Data Exploitation"
date: "28/01/2019"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(visNetwork)
library(DT)
```
```{r data, include=FALSE}
nodes_raw = read_csv(
  "got-nodes.csv",
  col_names = c("id", "label"),
  # set name of columns
  skip = 1 # skip column headers
  )
  
edges_raw = read_csv(
  "got-edges.csv",
  col_names = c("from", "to", "weight"),
  skip = 1
)
```

```{r preview, echo=FALSE}
visNetwork(nodes_raw,
           edges_raw,
           height = "500px",
           width = "100%")
```


[VisNetwork](https://datastorm-open.github.io/visNetwork/) is a powerful, flexible R library which very quickly allows you to create network diagrams such as the one shown above. This markdown gives an overview of the basics as we construct a network diagram from scratch, starting from the raw data.

## Network diagrams
### Loading the required libraries 

```{r eval = FALSE}
library(tidyverse)
library(visNetwork)

# if libraries are not installed:
# install.packages("visNetwork")
```

### The data

Creating network diagrams with VisNetwork requires two datasets:

* a **nodes** data.frame with an *id* column - a list of all unique nodes in the network, with their characteristics

* an **edges** data.frame *from* and *to* columns which link the *id* together - describes the relationships between the nodes

For our demonstration, we'll be using network data for character relationships within George R. R. Martin's A Song of Ice and Fire (also known as the HBO television adaptation **Game of Thrones**), specifically the third novel in the series. The nodes csv contains 107 different characters, and the edges csv contains 353 weighted relationships between those characters, which were calculated based on how many times two characters' names appeared within 15 words of one another in the novel.  

For more information, see here: https://github.com/melaniewalsh/sample-social-network-datasets/tree/master/sample-datasets/game-of-thrones

***

```{r nodes, echo=FALSE}
datatable(nodes_raw,
          caption = 'Table 1: Nodes dataset',
          class = 'cell-border stripe',
          rownames = FALSE,
          options = list(searching = FALSE)
)
```

***

```{r edges, echo = FALSE}
datatable(edges_raw,
          caption = 'Table 2: Edges dataset',
          class = 'cell-border stripe',
          rownames = FALSE
          )
```

***

### Our first network 

Upon reading these two datasets, we then load them into the visNetwork function as follows...

```{r first_network, message=FALSE, warning=FALSE}

# Load in the nodes data
nodes_raw = read_csv(
  "got-nodes.csv", # name of file saved in the working directory
  col_names = c("id", "label"), # set name of columns to conform to requirements
  skip = 1 # skip existing column headers
  )

# Load in the edges data
edges_raw = read_csv(
  "got-edges.csv",
  col_names = c("from", "to", "weight"),
  skip = 1
)

# Generate the default network diagram
visNetwork(nodes_raw,
           edges_raw,
           width = "100%")
```

### Customising the nodes

It's difficult to make sense of the network at present, how can we make this more useful? 

Let's highlight a few key characters by adding columns to the nodes dataset... 

```{r customising_nodes, message=FALSE, warning=FALSE}
nodes = nodes_raw %>% # pipe: pass in code *before* the pipe to the code *after* the pipe
  mutate(
    # Creating a column for the Stark, Lannister and Targaryen houses
    house = case_when(
    id %in% c("Arya", "Bran", "Jon", "Catelyn", "Robb", "Sansa") ~ "Stark",
    id %in% c("Tywin", "Tyrion", "Jaime", "Cersei") ~ "Lannister",
    id %in% c("Daenerys") ~ "Targaryen"
    ),
    # Colouring the nodes
    color = case_when(
    house == "Stark" ~ "slategray",
    house == "Lannister" ~ "gold",
    house == "Targaryen" ~ "red",
    TRUE ~ "lightgrey"
    ),
    # Sizing the nodes
    size = case_when(
    house %in% c("Stark", "Lannister", "Targaryen") ~ 40,
    TRUE ~ 10
    ),
    # Amending font size for Starks and Lannisters
    font.size = case_when(
    house %in% c("Stark", "Lannister", "Targaryen") ~ 50,
    TRUE ~ 15
  )
  )
```

```{r nodes_v2, echo=FALSE}
datatable(nodes,
          caption = 'Table 3: Nodes dataset with additional features',
          class = 'cell-border stripe',
          rownames = FALSE,
          options = list(searching = FALSE)
)
```

Let's read this enriched nodes dataset into the visNetwork function...

```{r network_v2, message=FALSE, warning=FALSE}
# Generate the diagram
visNetwork(nodes, # Reading in the new nodes dataset
           edges_raw,
           width = "100%")
```

### Customising the edges

We can also customise the edges in a similiar fashion. Recalling that the strength of the relationships between characters is found in the **weight** field, we can represent this with the width of the edges.

```{r edges2, message=FALSE, warning=FALSE}
edges = edges_raw %>%
  mutate(width = weight/5 + 1) # Scales down the default weight values, + 1 for a minimum value
```

```{r network_v3, message=FALSE, warning=FALSE}
# Generate the diagram
visNetwork(nodes,
           edges,
           width = "100%")
```

### Global options 

So far, we've been customising the network by adding new fields to our nodes and edges. There are a host of global options that can be set to the network as a whole. Let's add a few features to make the network more useful...

```{r}
visNetwork(nodes, edges, width = "100%") %>% 
  visOptions(highlightNearest = TRUE, # higlights nearest nodes upon selection
             nodesIdSelection = TRUE, # adds a dropdown for individual characters
             selectedBy = "house") %>% # adds a dropdown to select houses
  visExport(type = "png") # export functionality
```

## Hierarchical diagrams

With a few tweaks to the code and the data, visNetwork can also be used to construct network diagrams in a hierarchical layout.

Let's load in a different set of data and review...

```{r hierarchical_data, message=FALSE, warning=FALSE}

# Load in the nodes data
nodes_h = read_csv(
  "got-nodes-hierarchy.csv",
  col_names = c("id", "label"), # set name of columns to conform to requirements
  skip = 1 # skip existing column headers
  )

# Load in the edges data
edges_h = read_csv(
  "got-edges-hierarchy.csv",
  col_names = c("from", "to"), 
  skip = 1
)
```

***

```{r h_nodes_data_table, echo=FALSE}
datatable(nodes_h,
          caption = 'Table 4: Nodes hierarchical dataset',
          class = 'cell-border stripe',
          rownames = FALSE,
          options = list(searching = FALSE)
)
```

***

```{r h_edges_data_table, echo=FALSE}
datatable(edges_h,
          caption = 'Table 5: Edges hierarchical dataset',
          class = 'cell-border stripe',
          rownames = FALSE,
          options = list(searching = FALSE)
)
```

We can then construct the hierarchical diagram as follows...

```{r network_hierarchical, message=FALSE, warning=FALSE}
# Generate the diagram
visNetwork(nodes_h,
           edges_h,
           width = "100%") %>%
  visNodes(shape = "square") %>% # global options for shapes
  visEdges(arrows = "to") %>% # direction of arrows
  visHierarchicalLayout(sortMethod = "directed", # set hierarchical layout
                        nodeSpacing = 500) %>%
  visOptions(collapse = TRUE, manipulation = TRUE) # a few more options
```