---
title: "VisNetwork: an introduction"
author: "Faisal Samin, Data Exploitation"
date: "28/01/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(visNetwork)
library(DT)
```
```{r data, include=FALSE}
nodes = read_csv(
  "got-nodes.csv",
  col_names = c("id", "label"),
  # set name of columns
  skip = 1 # skip column headers
  )
  
edges = read_csv(
  "got-edges.csv",
  col_names = c("from", "to", "weight"),
  skip = 1
)
```
```{r final_network, echo=FALSE}
visNetwork(nodes,
         edges,
         height = "500px",
         width = "100%")
```


[VisNetwork](https://datastorm-open.github.io/visNetwork/) is a powerful, user-friendly R library which very quickly allows you to create network diagrams such as the one shown above. This markdown gives an overview of the basics as we construct the network above from scratch, starting from the raw data.

### Loading the required libraries 

```{r eval = FALSE}
library(tidyverse)
library(visNetwork)
```

### The data

Creating network diagrams with VisNetwork requires two datasets:

* a **nodes** data.frame with an *id* column - a list of all the nodes in the network, with their characteristics

* an **edges** data.frame *from* and *to* columns which link the *id* together - describes the relationships between the nodes

For our demonstration, we'll be using network data for character relationships within George R. R. Martin's A Song of Ice and Fire (also known as the HBO television adaptation **Game of Thrones**), specifically the third novel in the series. The nodes csv contains 107 different characters, and the edges csv contains 353 weighted relationships between those characters, which were calculated based on how many times two characters' names appeared within 15 words of one another in the novel.  

For more information, see here: https://github.com/melaniewalsh/sample-social-network-datasets/tree/master/sample-datasets/game-of-thrones

***

```{r nodes, echo=FALSE}
datatable(nodes,
          caption = 'Table 1: Nodes dataset',
          class = 'cell-border stripe',
          rownames = FALSE
)
```

***

```{r edges, echo = FALSE}
datatable(edges,
          caption = 'Table 2: Edges dataset',
          class = 'cell-border stripe',
          rownames = FALSE
          )
```

### Our first network 

We load in these two datasets and then load them into the visNetwork function.

```{r first_network, message=FALSE, warning=FALSE}

# Load in the nodes data
nodes = read_csv(
  "got-nodes.csv",
  col_names = c("id", "label"), # set name of columns to conform to requirements
  skip = 1 # skip existing column headers
  )

# Load in the edges data
edges = read_csv(
  "got-edges.csv",
  col_names = c("from", "to", "weight"),
  skip = 1
)

# Generate the default network diagram
visNetwork(nodes,
           edges,
           width = "100%")
```

### Customising the nodes

It's difficult to make sense of the network at present, it would help if we could add colours to represent characters. We can do this within the dataset. Let's highlight the two key houses, the Starks and the Lannisters. 

```{r network_colours, message=FALSE, warning=FALSE}
nodes2 = nodes %>%
  mutate(
    # Creating a column for the Stark and Lannister houses
    house = case_when(
    id %in% c("Arya", "Bran", "Jon", "Catelyn", "Robb", "Sansa", "Eddard") ~ "Stark",
    id %in% c("Tywin", "Tyrion", "Jaime", "Cersei") ~ "Lannister"
    ),
    # Colouring the nodes
    color = case_when(
    house == "Stark" ~ "slategray",
    house == "Lannister" ~ "firebrick",
    TRUE ~ "lightgrey"
    ),
    # Sizing the nodes
    size = case_when(
    house %in% c("Stark", "Lannister") ~ 40,
    TRUE ~ 10
    ),
    # Amending font size for Starks and Lannisters
    font.size = case_when(
    house %in% c("Stark", "Lannister") ~ 50,
    TRUE ~ 15
  )
  )
```

```{r nodes2, echo=FALSE}
datatable(nodes2,
          caption = 'Table 3: Nodes dataset with additional feautures',
          class = 'cell-border stripe',
          rownames = FALSE,
          options = list(searching = FALSE)
)
```

```{r network_v2, message=FALSE, warning=FALSE}
# Generate the diagram
visNetwork(nodes2,
           edges,
           width = "100%")
```

### Customising the edges

We can also customise the edges in a similiar fashion. Recalling that the strength of the relationships between characters is found in the **weight** field, we can represent this with the width of the edges.

```{r edges2, message=FALSE, warning=FALSE}
edges2 = edges %>%
  mutate(width = weight/5 + 1)
```

```{r network_v3, message=FALSE, warning=FALSE}
# Generate the diagram
visNetwork(nodes2,
           edges2,
           width = "100%")
```

### Global options 

Finally, we can set a few useful global options from within the visNetwork function.

```{r}
visNetwork(nodes2, edges2, width = "100%") %>% 
  visNodes(shapeProperties = list(useBorderWithImage = TRUE)) %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE,
             selectedBy = "house"
  ) %>%
  visExport(type = "png")
```

